---
title: "Hyperparameter tuning in abm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{v00_hyperparameter_tuning}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r dependencies, include=FALSE}
library(knitr)
# devtools::install_github('sjevelazco/adm')
```

### Introduction

Using multiple algorithms and performing hyperparameters tuning is crucial for diversifying species abundance modeling.
*adm* facilitates modeling, hyperparameter tuning, and validation of nine algorithms.
Functions with the *tune_abund* prefix allow users to perform model tuning by using an array of model-specific hyperparameters.
In *adm*, tuning is performed using a grid-search approach, i.e., by evaluating model performance under an array of possible hyperparameter combinations.
To implement this approach, the user provides a data frame with hyperparameters as columns and the hyperparameter values to be tested as rows.

| Abbreviation | Taxonomic database                                        |
|--------------|-----------------------------------------------------------|
| col          | Catalogue of Life                                         |
| fb           | FishBase                                                  |
| itis         | Integrated Taxonomic Information System                   |
| iucn         | International Union for Conservation of Nature's Red List |
| gbif         | Global Biodiversity Information Facility                  |
| ncbi         | National Center for Biotechnology Information             |
| ott          | OpenTree Taxonomy                                         |
| tpl          | The Plant List                                            |
| slb          | SeaLifeBase                                               |
| wd           | Wikidata                                                  |
